---
title: "DS605-HW12"
author: "George Cruz"
date: "4/25/2021"
output: 
  rmdformats::material:
  code_folding: hide
  #prettydoc::html_pretty:
  #  theme: architect
   # highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
```

# Introduction to Data 

The attached `who.csv` dataset contains real-world data from 2008. The variables included follow.

**Country:** name of the country\
**LifeExp:** average life expectancy for the country in years\
**InfantSurvival:** proportion of those surviving to one year or more\
**Under5Survival:** proportion of those surviving to five years or more\
**TBFree:** proportion of the population without TB.\
**PropMD:** proportion of the population who are MDs\
**PropRN:** proportion of the population who are RNs\
**PersExp:** mean personal expenditures on healthcare in US dollars at average exchange rate\
**GovtExp:** mean government expenditures per capita on healthcare, US dollars at average exchange rate\
**TotExp:** sum of personal and government expenditures.\

```{r}
who <- read.csv("https://raw.githubusercontent.com/georg4re/DS605/main/data/who.csv")
summary(who)
```

# 1. ScatterPlot and Initial Linear Regression
Provide a scatterplot of LifeExp~TotExp, and run simple linear regression. Do not transform the variables. Provide and interpret the F statistics, R^2, standard error,and p-values only. Discuss whether the assumptions of simple linear regression met.

```{r}
ggplot(who, aes(x = TotExp, y = LifeExp)) + 
  geom_point() + 
  geom_smooth(method='lm', formula= y~x)
```

```{r}
le_te_lm <- lm(LifeExp ~ TotExp, data = who)
summary(le_te_lm)
```

The `F-statistic` is *65.26* with a p-value close to 0 gives us an indication of a significant relationship between Total Expenditure and Life Expectancy.\
The $R^2$ tells us that this model accounts for only **25.77%** of the variation of the data.

```{r, echo=FALSE}

  par(mfrow=c(2,2))

  #Linearity
  plot(fitted(le_te_lm),
     resid(le_te_lm),
     main = "Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals",
     col = '#009587')
  abline(h=0, lty = 2)

  #Normality of Residuals
  qqnorm(resid(le_te_lm), 
       col = '#009587')
  qqline(resid(le_te_lm), 
       lty = 2)

  #Variablity
  hist(resid(le_te_lm), 
     breaks=30, 
     prob=TRUE,
     main = "Histogram: Residuals", 
     xlab = "Residuals", 
     ylab = "Density",
     col = '#009587')
  curve(dnorm(x, 
            mean=mean(resid(le_te_lm)), 
            sd=sd(resid(le_te_lm))), 
      lty = 2, 
      add = TRUE)

  #Independence of Errors
  plot(resid(le_te_lm), 
     main="Residuals",
     ylab = "Residuals",
     col = '#009587')
  abline(h = 0, 
       lty = 2)
  abline(h = sd(resid(le_te_lm)),
         lty = 3)
  abline(h = -sd(resid(le_te_lm)),
         lty = 3)
```

**Are the assumptions met?**

1. The residuals appear to be randomly distributed on the plot.\
2. Most of the residuals fall on the theoretical normal line.\
3. According to the histogram, the distribution of the residuals is slightly left skewed short long tails.\
4. The variation of the residuals appears to be constant accross the model and the residuals are independent from each other.


# 2. Which Model is Better? 
Raise life expectancy to the 4.6 power (i.e., LifeExp^4.6).\
Raise total expenditures to the 0.06 power (nearly a log transform, TotExp^.06).\

```{r}
who_2 <- who %>%
  mutate(TotExp = TotExp^0.06,
         LifeExp = LifeExp^4.6)
```

Plot LifeExp^4.6 as a function of TotExp^.06, and re-run the simple regression model using the transformed variables.\

```{r}
ggplot(who_2, aes(x = TotExp, y = LifeExp)) + 
  geom_point() + 
  geom_smooth(method='lm', formula= y~x)
```

```{r, echo=FALSE}
model2 <- lm(LifeExp ~ TotExp, data = who_2)
summary(model2)
```

```{r, echo=FALSE}

  par(mfrow=c(2,2))

  #Linearity
  plot(fitted(model2),
     resid(model2),
     main = "Residuals vs Fitted", 
     xlab = "Fitted Values", 
     ylab = "Residuals",
     col = '#009587')
  abline(h=0, lty = 2)

  #Normality of Residuals
  qqnorm(resid(model2), 
       col = '#009587')
  qqline(resid(model2), 
       lty = 2)

  #Variablity
  hist(resid(model2), 
     breaks=30, 
     prob=TRUE,
     main = "Histogram: Residuals", 
     xlab = "Residuals", 
     ylab = "Density",
     col = '#009587')
  curve(dnorm(x, 
            mean=mean(resid(model2)), 
            sd=sd(resid(model2))), 
      lty = 2, 
      add = TRUE)

  #Independence of Errors
  plot(resid(model2), 
     main="Residuals",
     ylab = "Residuals",
     col = '#009587')
  abline(h = 0, 
       lty = 2)
  abline(h = sd(resid(model2)),
         lty = 3)
  abline(h = -sd(resid(model2)),
         lty = 3)

```

**Provide and interpret the F statistics, R^2, standard error, and p-values.**\

- **F Statistics & P-value:** the value is *507.7* with p-value very close to zero, which means there is significant evidence that the current model is better than the initial model. If we recall that the original model's `F-statistic` was *65.26*, we can assume this is a better model.\
- **$R^2$:** around 73% of the variability of the data can be explained by the current model vs 26% with the initial one.\
- **Standard error:** the standard error is way larger than the initial model's.  Not entirely sure what that means\


**Which model is "better?"**

The second model is a better one.  It accounts for 73% of the variability.  The distribution of the residuals is nearly normal and independent.


# 3. Forecast 
Using the results from 3, forecast life expectancy when $TotExp^.06 =1.5$. 

```{r}
at_1.5 <- (-736527910 + 620060216 * 1.5) ^ (1/4.6)

paste("Life Expectancy when Tot Exp = 1.5 is: ", at_1.5)
```

Then forecast life expectancy when $TotExp^.06=2.5$

```{r}
at_2.5 <- (-736527910 + 620060216 * 2.5) ^ (1/4.6)

paste("Life Expectancy when Tot Exp = 2.5 is: ", at_2.5)
```


# 4. Build Models.
Build the following multiple regression model and interpret the F Statistics, R^2, standard error, and p-values. How good is the model?
$LifeExp = b_0+b_1 * PropMd + b_2 * TotExp + b_3 * PropMD * TotExp$

```{r}
model3 <- lm(LifeExp ~ PropMD + TotExp + PropMD*TotExp , who)
summary(model3)
```

- **F Statistics & P-value:** the value is *34.49* with p-value close to zero, the `F-statistic` is lower than the original model's which was *65.26*.\
- **$R^2$:** around 36% of the variability of the data can be explained by the current model vs 26% with the initial one.\
- **Standard error:** the residual standard error is 8.765, smaller than the initial model's.\

Overall, this model seems better than the initial one but not as good as the transformed model.


# 5. Forecast LifeExp 
when PropMD=.03 and TotExp = 14. Does this forecast seem realistic? Why or why not?

```{r}
62.77 + (1497 * 0.03) + (0.00007233 * 14) - (0.006026 * (0.03 * 14))
```

We saw that as the Total Expenditure increases, so does the life expectancy.  That said, 107 appears to be a little high and unrealistic.  Only two countries have a PropMD of 0.03 or higher and their TotExp are way higher than 14.  The fact that these countries have LifeExp around 80 seem to imply that this proyection is totally unrealistic. 
